Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-11-03 manu-draft.qmd
minor results edits

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$. Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.  The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).  The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.  The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay (@fig-meteowqraw).  Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).  Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.  As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]).  This variation may also be related to differences among the monitoring stations in each segment.  Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).  No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).  However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.  Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).

>>>>

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$. Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement). Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrntab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends. The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022. The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay (@fig-meteowqraw). Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab). Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present. As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]). This variation may also be related to differences among the monitoring stations in each segment as the averages in @fig-meteowqraw combine monthly and spatial variation. Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab). No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement). However, the EPC dataset showed significant increasing temperature across all time periods and bay segments. Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).
